---
title: "Project 3"
author: "Richard Affolter"
date: "11 3 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Problem  6: Hidden Markov Models

According to slide 18 the HMM is parametrized by 

- The initial state probabilities, which can take up K different values, but since they have to sum up to 1 (they are probabilities) we have K-1 degrees of freedom.

- The transition probabilities, where we have aa K$\cdot$K matrix, but since total the probability of jumping from a state i into a state j must sum up to 1,  we have K$\cdot$(K-1) degrees of freedom.

- The emission probabilities, where we have K$\cdot$M matrix, but since the probability of emitting something (from any state i) must sum up to 1 we have K$\cdot$(M-1) degrees of freedom.

In total we have K-1 + K$\cdot$(K-1) + K$\cdot$(M-1) = K$\cdot$(K+M-1)-1 possible unique parameters.


## Problem 7: Predicting protein secondary structure using HMMs

```{r load libraries, warning=FALSE, message=FALSE}
library(dplyr)
```


### (a) Read `proteins_train.tsv`, `proteins_test.tsv` and `proteins_new.tsv` into the memory and store each in a data.frame

```{r, load data}
p_train = read.table(file = 'proteins_train.tsv', sep = '\t', header = FALSE)
p_test = read.table(file = 'proteins_test.tsv', sep = '\t', header = FALSE)
p_new = read.table(file = 'proteins_new.tsv', sep = '\t', header = FALSE)
```


### (b) Estimate the vector of initial state probabilities $I$, the matrix of transition probabilities $T$ and the matrix for emission probabilities $E$.

```{r, copy ss and aa}

# copy section from viterbi.r
unique.ss <- c("B", "C", "E", "G", "H", "I", "S", "T")
unique.aa <- c("A", "C", "D", "E", "F", "G", "H", "I",
               "K", "L", "M", "N", "P", "Q", "R", "S",
               "T", "U", "V", "W", "X", "Y")
```
```{r, calculate inital states}
get_I = function(df){
  I = rep(0, length(unique.ss))
  names(I) = unique.ss
  
  for(i in 1:dim(p_train)[1]){
    letter = substr(df[i,3],1,1)
    I[letter] = I[letter] + 1
  }

  return(I = I/sum(I))
}
I = get_I(p_train)
```

```{r, calculate transition}
get_T = function(df){
  T = matrix(0,nrow = length(unique.ss), ncol = length(unique.ss))
  rownames(T) = unique.ss
  colnames(T) = unique.ss
  
  for(i in 1:dim(df)[1]){
    sec_struc = strsplit(df[i,3], split = "") %>% unlist()
    for(j in 1:(length(sec_struc)-1)){
      T[sec_struc[j],sec_struc[j+1]] = T[sec_struc[j],sec_struc[j+1]] + 1
    }
  }
  return(T = T/rowSums(T))
}
T = get_T(p_train) 
```

```{r, calculate emmission}
get_E = function(df){
  E = matrix(0,nrow = length(unique.ss), ncol = length(unique.aa))
  rownames(E) = unique.ss
  colnames(E) = unique.aa
  
  for(i in 1:dim(df)[1]){
    sec_struc = strsplit(df[i,3], split = "") %>% unlist()
    aa_seq = strsplit(df[i,2], split = "") %>% unlist()
    for(j in 1:length(sec_struc)){
      E[sec_struc[j],aa_seq[j]] = E[sec_struc[j],aa_seq[j]] + 1
    }
  }
  return(E = E/rowSums(E))
}
E = get_E(p_train)
```
```{r print EIT}
print(E)
print(T)
print(I)
```


### (c)Estimate the stationary distribution $\pi$ of the Markov chain.

We know from slide 9 that $\pi$ is the solution of $\pi^t = \pi^tT$.
If we transpose we get an eigenvalue problem:
$$
(\pi^t)^\intercal = (\pi^tT)^\intercal = T^\intercal(\pi^t)^\intercal
$$
So we are looking for the eigenvector of $T^\intercal$ with eigenvalue 1.
```{r, calculate stationary distribution, warning=FALSE, message=FALSE}
library(dplyr)
ev = eigen(t(T))
# find the eigenvalue with value 1
pos = which(near(ev$values,1))

#normalize
Pi = ev$vectors[,pos]/sum(ev$vectors[,pos])
names(Pi) = unique.ss
```
```{r, print Pi}
print(Pi)
```


### (d) Predict the latent state sequence $Z$ of a protein’s amino acid sequence $X$ using the Viterbi algorithm

```{r, run the viterbi for the 3 data sets}
source("viterbi.r")
get_prediction = function(x){
  df = data.frame(AminoAcids = x$V2)
  tmp = viterbi(E=log(E), Tr = log(T), I=log(I),p=df)
  return(tmp$PredictedStructure)
}

p_train$PredictedStructure = get_prediction(p_train)
p_test$PredictedStructure = get_prediction(p_test)
p_new$PredictedStructure = get_prediction(p_new)
```
```{r save p_new as .tsv}
#  save p_new as a new tsv
write.table(p_new, file = "proteins_new.tsv", sep = "\t",
            col.names = FALSE, row.names = FALSE)
```


### (e) Estimate confidence intervals for $I$, $E$ and $T$ with bootstrapping


```{r, bootstrap, eval=TRUE}
# initialize lists to store the intermediate results of each bootstrap
I_list = list()
T_list = list()
E_list = list()

num_row = dim(p_train)[1]
num_boot = 1000
set.seed(123)
# resample the data and get T,E,I
for(i in 1:num_boot){
  resample_idx = sample(num_row,num_row, replace = TRUE)
  resampled_data = p_train[resample_idx,]
  I_list[[i]] = get_I(resampled_data)
  T_list[[i]] = get_T(resampled_data)
  E_list[[i]] = get_E(resampled_data)
}

# transform to 3D array
T_simp = simplify2array(T_list)
E_simp = simplify2array(E_list)
I_simp = simplify2array(I_list)
```

```{r, get CI for each entry}
# get CI for T
T_simp[is.na(T_simp)] = 0 # set Na to 0
dim1 = dim(T_simp)[1]
dim2 = dim(T_simp)[2]
T_CI = array(rep(0,dim1*dim2*2),dim = c(dim1,dim2,2),
             dimnames = list(unique.ss, unique.ss,
                             c("lower_bound", "upper_bound")))

for(i in 1:dim1){
  for(j in 1:dim2){
    T_simp[i,j,] %>% quantile(probs = c(0.025, 0.975))-> CI
    T_CI[i,j,] = CI
  }
}


# get CI for E
E_simp[is.na(E_simp)] = 0 # set Na to 0
dim1 = dim(E_simp)[1]
dim2 = dim(E_simp)[2]
E_CI = array(rep(0,dim1*dim2*2),dim = c(dim1,dim2,2),
             dimnames = list(unique.ss, unique.aa,
                             c("lower_bound", "upper_bound")))

for(i in 1:dim1){
  for(j in 1:dim2){
    E_simp[i,j,] %>% quantile(probs = c(0.025, 0.975))-> CI
    E_CI[i,j,] = CI
  }
}

# get CI for I
I_simp[is.na(I_simp)] = 0 # set Na to 0
dim1 = dim(I_simp)[1]
I_CI = array(rep(0,dim1*2),dim = c(dim1,2),
             dimnames = list(unique.ss,
                             c("lower_bound", "upper_bound")))

for(i in 1:dim1){
  I_simp[i,] %>% quantile(probs = c(0.025, 0.975))-> CI
  I_CI[i,] = CI
}
```
```{r, print the CIs}
print(E_CI)
print(T_CI)
print(I_CI)
```


### (f) Compute the accuracy of the predicted secondary structure for the data.frame of `proteins_test.tsv`

```{r, calculate accuracies}
accuracies = rep(0, nrow(p_test))
for(i in 1:nrow(p_test)){
  for(j in 1:nchar(p_test$V3[i])){
    if(substr(p_test$V3[i],j,j) == substr(p_test$PredictedStructure[i],j,j)){
      accuracies[i] = accuracies[i]+1
    }
  } 
  accuracies[i] = accuracies[i]/nchar(p_test$V3[i])
}
summary(accuracies)
```
The mean accuracy over all sequences is `r mean(accuracies)`.

### (g) Randomly guess secondary structures and compare to the Viterbi

For the probability of the random guessing the stationary distribution is used.
```{r, randomly guess the accuracies}
random_accuracies = rep(0, nrow(p_test))
for(i in 1:nrow(p_test)){
  random_structure = sample(unique.ss, size=nchar(p_test$V3[i]),
                            replace = TRUE,prob = Pi)
  random_structure = paste(random_structure, collapse = "")
  for(j in 1:nchar(p_test$V3[i])){
    if(substr(p_test$V3[i],j,j) == substr(random_structure,j,j)){
      random_accuracies[i] = random_accuracies[i]+1
    }
  } 
  random_accuracies[i] = random_accuracies[i]/nchar(p_test$V3[i])
}
summary(random_accuracies)
```
The mean accuracy of the random sequences over all sequences is `r mean(random_accuracies)`.

```{r, plot the accuracies}
par(mfrow=c(1,2))
boxplot(accuracies, main = "Viterbi prediction", ylim = c(0, 1))
boxplot(random_accuracies, main = "Random prediction", ylim = c(0,1))
```
We can see that the Viterbi algorithm performs better than random guessing.